A thermodynamic view of dusty protoplanetary disks: axisymmetric stability

### 2016-10-22 _region_.tex
head to toe edits

\section{
  Dusty vertical shear instability}\label{results}

We now present numerical solutions to the linearized equations in a 
vertically stratified disk. We consider perfectly coupled dust 
($\tstop=0$) so that the equilibria defined in \S\ref{eqm} are exact
steady states. In this limit any growing modes must be associated with
the imposed temperature gradient (\S\ref{limits}). We thus 
examine the effect of dust-loading on the VSI, previously studied in 
pure gas disks \citep[][\citetalias{lin15} in this section]{lin15}.

@@

>>>>

\message{ !name(result.tex)}
\message{ !name(result.tex) !offset(-2) }
\section{
  Dusty vertical shear instability}\label{results}

We present numerical solutions for  
vertically stratified disks assuming perfectly coupled dust 
($\tstop=0$). This is so that the equilibria defined in \S\ref{eqm} are exact
steady states. In this limit any growing modes must be associated with
the imposed temperature gradient (\S\ref{limits}). We are thus  
examining the effect of dust-loading on the VSI, previously studied in 
pure gas disks \citep[][\citetalias{lin15} in this section]{lin15}.

@@

\message{ !name(result.tex) !offset(-360) }

### 2016-10-22 appendix.tex
fixed some typos, corrected kappa^2 dependence on dust
head to toe edits

\section{Linearized dust diffusion}\label{lin_dust}
We consider small grains in the Epstein regime, with fixed internal
density and size, so that
\begin{align}
  \tstop  =  \frac{K}{\rho c_s}
\end{align}
\citep{price15}, where $K$ is a constant. Then the dust diffusion
function becomes
\begin{align}
  \mathcal{C} \equiv c_s^2\nabla\cdot\left(\tepsilon\tstop\nabla
  P\right) = - K c_s^2 \nabla\cdot
  \left(\frac{\tepsilon}{c_s}\bm{F}\right) =
  -Kc_s \left(\bm{F}\cdot\nabla\tepsilon + \tepsilon \nabla\cdot\bm{F}
  - \frac{1}{2}\tepsilon\bm{F}\cdot\nabla\ln{c_s^2}\right).  
\end{align}
Linearizing,
\begin{align}
  - \frac{\dd\mathcal{C}}{Kc_s} = \bm{F}\cdot\nabla\dd\tepsilon +
  \dd\bm{F}\cdot\nabla\tepsilon + \dd\tepsilon\nabla\cdot\bm{F} +
  \tepsilon\nabla\cdot\dd\bm{F} -
  \frac{1}{2}\nabla\ln{c_s^2}\cdot\left(\bm{F}\dd\tepsilon + \tepsilon
  \dd\bm{F}\right). 
\end{align}
The linearized dust-fraction $\dd\tepsilon$ and its derivatives in the
radially-local approximation are given by
\begin{align}
  \dd\tepsilon      &= (1-\tepsilon)W - \frac{Q}{c_s^2}, \\
  \p_r\dd\tepsilon &= \left[(1-\tepsilon)\left(\ii k_x -
    \p_r\ln{\rho}\right) - \p_r\tepsilon\right]W + \left(\p_r\ln{c_s^2}
  + \p_r\ln{\rho} - \ii k_x \right)\frac{Q}{c_s^2},\\
  \dd\tepsilon^\prime &= (1-\tepsilon)W^\prime - \tepsilon^\prime W -
  \left(\frac{Q}{c_s^2}\right)^\prime. 
\end{align}

@@

\section{One-fluid dispersion relation for the streaming instability}\label{compressible_streaming}
We consider an unstratified disk with $\Phi = \Phi (r)$ so that $\p_zP = \p_z\rho =
0$. The background $\p_rP/\rho$ and $\tepsilon$ are constants. 
We Fourier analyze in $r$ and $z$ so that $\p_z\to \ii k_z$ and
$\p_r\to \ii k_x$ when acting on perturbations, and denote $|\bm{k}|^2
\equiv k_x^2 + k_z^2$. We neglect 
background gradients when compared with that of perturbations. The linearized
equations are, after eliminating the azimuthal velocity:

\begin{align}
  \ii \sigma W &=\nabla\cdot \dd \bm{ v} = \ii k_x \dd v_x + \ii k_z \dd v_z,\\
    \sigma^2 \dd v_x &= \kappa^2 \dd v_x - \ii \sigma F_r W +
    k_x\sigma Q,\\
  -\ii\sigma\dd v_z &= -\ii k_zQ,\\
\ii \zeta \sigma  Q & = \frac{P}{\rho} \nabla \cdot \dd\bm{v}   -
  \frac{\dd \mathcal{C}}{\rho}. 
\end{align}
The artifical factor $\zeta = 1$  is inserted to keep track of the
left-hand-side of the energy equation. We note that the incompressible
condition used by \cite{jacquet11} is obtained by setting $\zeta$ to zero. 
The linearized dust diffusion
function (Appendix \ref{lin_dust}), 
under the above approximations, is 
\begin{align}
-\frac{\dd \mathcal{C}}{\rho}  = \tstop c_s^2 \left[\ii k_x F_r\left( 1-2\tepsilon\right)
  W + \tepsilon |\bm{k}|^2 Q\right]. 
\end{align}

>>>>

\section{Linearized dust diffusion}\label{lin_dust}
We consider small grains in the Epstein regime, with fixed internal
density and size, so that
\begin{align}
  \tstop  =  \frac{K}{\rho c_s}\label{epstein}
\end{align}
\citep{price15}, where $K$ is a constant. Then the dust diffusion
function becomes
\begin{align}
  \mathcal{C} \equiv c_s^2\nabla\cdot\left(\tepsilon\tstop\nabla
  P\right) = - K c_s^2 \nabla\cdot
  \left(\frac{\tepsilon}{c_s}\bm{F}\right) =
  -Kc_s \left(\bm{F}\cdot\nabla\tepsilon + \tepsilon \nabla\cdot\bm{F}
  - \frac{1}{2}\tepsilon\bm{F}\cdot\nabla\ln{c_s^2}\right).  
\end{align}
Linearizing,
\begin{align}
  - \frac{\dd\mathcal{C}}{Kc_s} = \bm{F}\cdot\nabla\dd\tepsilon +
  \dd\bm{F}\cdot\nabla\tepsilon + \dd\tepsilon\nabla\cdot\bm{F} +
  \tepsilon\nabla\cdot\dd\bm{F} -
  \frac{1}{2}\nabla\ln{c_s^2}\cdot\left(\bm{F}\dd\tepsilon + \tepsilon
  \dd\bm{F}\right). 
\end{align}
The linearized dust-fraction $\dd\tepsilon$ and its derivatives in the
radially-local approximation are given by
\begin{align}
  \dd\tepsilon      &= (1-\tepsilon)W - \frac{Q}{c_s^2}, \\
  \p_r\dd\tepsilon &= \left[(1-\tepsilon)\left(\ii k_x -
    \p_r\ln{\rho}\right) - \p_r\tepsilon\right]W + \left(\p_r\ln{c_s^2}
  + \p_r\ln{\rho} - \ii k_x \right)\frac{Q}{c_s^2},\\
  \dd\tepsilon^\prime &= (1-\tepsilon)W^\prime - \tepsilon^\prime W -
  \left(\frac{Q}{c_s^2}\right)^\prime. 
\end{align}

@@

\section{One-fluid dispersion relation for the streaming instability}\label{compressible_streaming}
We consider an unstratified disk with $\Phi = \Phi (r)$ so that $\p_zP = \p_z\rho =
0$. The background $\p_rP/\rho$, $\tepsilon$ and $c_s^2$ are constant
input parameters. Then for the Epstein drag law, Eq. \ref{epstein}, we
have $\mathcal{C}=0$ in the background state. 
We Fourier analyze in $r$ and $z$ so that $\p_z\to \ii k_z$ and
$\p_r\to \ii k_x$ when acting on perturbations, and denote $|\bm{k}|^2
\equiv k_x^2 + k_z^2$. We neglect 
background gradients when compared with that of
perturbations. The linearized
equations are, after eliminating the azimuthal velocity:

\begin{align}
  \ii \sigma W &=\nabla\cdot \dd \bm{ v} = \ii k_x \dd v_x + \ii k_z \dd v_z,\label{streaming_mass}\\
    \sigma^2 \dd v_x &= \kappa^2 \dd v_x - \ii \sigma F_r W +
    k_x\sigma Q,\label{streaming_vx}\\
  -\ii\sigma\dd v_z &= -\ii k_zQ,\label{streaming_vz}\\
\ii \zeta \sigma  Q & = \frac{P}{\rho} \nabla \cdot \dd\bm{v}   -
  \frac{\dd \mathcal{C}}{\rho}. 
\end{align}
%{\bf bg pressure gradient neglected in energy eq (is this ok? could
%  fix, but get more complicated dispersion}
The artifical factor $\zeta = 1$  is inserted to keep track of the
left-hand-side of the energy equation. We note that the incompressible
condition used by \cite{jacquet11} is obtained by setting $\zeta$ to zero. 
The linearized dust diffusion
function (Appendix \ref{lin_dust}), 
under the above approximations, is 
\begin{align}
-\frac{\dd \mathcal{C}}{\rho}  = \tstop c_s^2 \left[\ii k_x F_r\left( 1-2\tepsilon\right)
  W + \tepsilon |\bm{k}|^2 Q\right]. 
\end{align}
%{\bf only the second term contributes to work done? dont think so}

### 2016-10-22 linear.tex
last update until november, for real
head to toe edits

\section{Compressible streaming instability}\label{si}
In the special case of an unstratified disk (neglecting the vertical
component of the stellar gravity), we may also Fourier analyze in $z$
to obtain algebraic dispersion relation of the form  
$\sum_{j=0}^{5}c_j(k_x,k_z)\sigma^j = 0$, where $k_z$ is a real
vertical wavenumber. The coefficients $c_j$ can be read 
off Eq. \ref{streaming_dispersion} in  Appendix \ref{compressible_streaming}. 
There we also show that this dispersion relation reduces to that for
the streaming instability (SI) in the limit of incompressible gas and small
$\tstop$ \citep{youdin05a,jacquet11}.

{\bf note: kappa2 accounts for dust effect assuming smallh=0.05}
We solve Eq. \ref{streaming_dispersion} for selected cases where SI  
growth rate have been 
calculated analytically from full two-fluid disk models and measured 
from direct numerical particle-gas simulations 
\citep[namely][]{youdin07b,bai10b}. These authors use normalized
wavenumbers $K_x = \eta r k_x$ and similarly for $K_z$, where
\begin{align} 
  \eta \equiv -\frac{1}{2\rhog r\OmK^2}\frac{\p P}{\p r} = 
  \frac{1}{2\left(1-\tepsilon\right)}\frac{F_r}{r\OmK^2}, 
\end{align} 
measures the pressure offset of Keplerian rotation. We use
$\eta=0.05c_s/r\OmK$. They also define the stopping time as 
$\tau_\mathrm{s}=\tstop/(1-\tepsilon)$.

@@

We also checked the growth rates $s$ satisfy 
\begin{align} 
  s = \frac{\left|\sigma\right|^2\imag\left(\Delta P
    \Delta\rho^*/\rho\right)}{2\real\left(\sigma\right)\mathcal{I}^2}, \label{si_check}
\end{align}
as implied by Eq. \ref{thermal_instability} and
Eq. \ref{pdv}. Interestingly, we find that for cases with
$\epsilon>1$, Eq. \ref{si_check} can be satisified with $\Delta P \to
\Delta P - \dd P$, i.e. the advection part of the Lagragian
pressure perturbation is dominant. Conversely, for `linB' with $\epsilon
< 1$, one can take $\Delta P \simeq \dd P$ in Eq. \ref{si_check}. This
suggests that for SI in gas-dominated disks the distinction between
Eulerian and Lagragian presure perturbations is unimportant.

{\bf note: important to use largrangian pressure pert}
In Table \ref{si_compare} we also calculate the phase difference
between the Lagragian pressure perturbation and density perturbations
as    
\begin{align*} 
\varphi = -\sgn(\omega)\arg\left(\Delta P\Delta\rho^*\right), 
\end{align*}
so that $\varphi > 0 $ indicates pressure lagging behind density, which
is true for all the cases. In particular, `linB' indicates that growth
rates vanish as $\varphi \to 0$.  
Thus the streaming instability stems  
from the gas pressure evolution lagging behind that of the total
density (and hence behind the dust density for isothermal gas).

@@

>>>>

\section{Compressible streaming instability}\label{si}
In the special case of an unstratified disk (neglecting the vertical
component of the stellar gravity), we may also Fourier analyze in $z$
to obtain an algebraic dispersion relation of the form  
$\sum_{j=0}^{5}c_j(k_x,k_z)\sigma^j = 0$, where $k_z$ is a real
vertical wavenumber. The coefficients $c_j$ can be read 
off Eq. \ref{streaming_dispersion} in  Appendix \ref{compressible_streaming}. 
There we also show that this dispersion relation reduces to that for
the streaming instability (SI) in the limit of incompressible gas and small
$\tstop$ \citep{youdin05a,jacquet11}.

{\bf note: kappa2 accounts for dust effect assuming smallh=0.05}
We solve Eq. \ref{streaming_dispersion} for selected cases where SI  
growth rates have been 
calculated analytically from full two-fluid disk models and measured 
from direct numerical particle-gas simulations 
\citep[namely][]{youdin07b,bai10b}. These authors use normalized
wavenumbers $K_x = \eta r k_x$ and similarly for $K_z$, where
\begin{align} 
  \eta \equiv -\frac{1}{2\rhog r\OmK^2}\frac{\p P}{\p r} = 
  \frac{1}{2\left(1-\tepsilon\right)}\frac{F_r}{r\OmK^2}, 
\end{align} 
measures the pressure offset of Keplerian rotation. We use
$\eta=0.05c_s/r\OmK$. They also define the stopping time as 
$\tau_\mathrm{s}=\tstop/(1-\tepsilon)$.

@@

{\bf note: important to use largrangian pressure pert properly}
We also checked the growth rates $s$ satisfy 
\begin{align} 
  s = \frac{\left|\sigma\right|^2\imag\left(\Delta P
    \Delta\rho^*/\rho\right)}{2\real\left(\sigma\right)\rho\left(\left|\dd
  v_x\right|^2+\left|\dd
  v_z\right|^2\right)}, \label{si_check}
\end{align}
as implied by Eq. \ref{thermal_instability} and
Eq. \ref{pdv}. %, or directly from
%Eq. \ref{streaming_mass}---\ref{streaming_vz}. 
Interestingly, we find that for cases with
$\epsilon>1$, Eq. \ref{si_check} can be satisified with $\Delta
P\simeq \ii\dd v_x \p_rP/\sigma$, i.e. the radial pressure gradient
is responsible for growth. Conversely, for `linB' with $\epsilon
< 1$, one can approximate $\Delta P \simeq \dd P$ in Eq. \ref{si_check}. This
suggests that for SI in gas-dominated disks, the distinction between
Eulerian and Lagragian pressure perturbations is unimportant. 
{\bf but weak growth in this case}
%However
%in that case the gro

In Table \ref{si_compare} we also calculate the phase difference
between the Lagragian pressure perturbation and density perturbations
as    
\begin{align*} 
\varphi \equiv -\sgn(\omega)\arg\left(\Delta P\Delta\rho^*\right)
        = \sgn(\omega)\arg\left(\Delta\rhog^*\Delta\rhod\right),\notag
\end{align*}
%where we have applied the isothermal equation of state for the second equality.
then $\varphi > 0 $ indicates pressure lagging behind density, which
is true for all the cases. In particular, `linB' indicates that growth
rates vanish as $\varphi \to 0$.  
Thus the streaming instability indeed stems  
from the gas pressure evolution lagging behind that of the dust 
density.

@@

{\bf suppose large scale pressure grad drops outwards. 
pert the sys by kicking more dust inwards. gas moves out, into region 
  of lower pressure in the bg. i.e. create pressure bump at larger
  radius. it attracts the first dust particles back out, plus some more
  because bump is larger. (move gas from higher density to lower
  density produce asymmetric bump/trough.}

### 2016-10-22 model.tex
head to toe edits

For a given distribution of the dust-fraction $\tepsilon$ (or
dust-to-gas ratio $\epsilon$), the 
mass and momentum Eqs. \ref{masseq}---\ref{momeq} admit     
axisymmetric steady states with $\rho(r,z)$ and 
$\bm{v}=r\Omega(r,z)\hat{\bm{\phi}}$ where $\Omega = v_\phi/r$, which satisfy 
\begin{align}
  r\Omega^2 &= \frac{\p \Phi}{\p r} + \frac{1}{\rho}\frac{\p P}{\p
    r},\label{steady_momr}\\
  0 & = \frac{\p\Phi}{\p z} + \frac{1}{\rho}\frac{\p P}{\p z},\label{steady_momz}
%  0 & = \nabla\cdot\left(\tepsilon\tstop\nabla P\right) \label{steady_dust}
\end{align}
with $P=P(\tepsilon,\rho)$ given by the equation of state
(Eq. \ref{eos}). An explicit solution is presented in
\S\ref{steady_state}.  
However, disk structures obtained this way generally do not satisfy 
the steady state effective energy Eq. \ref{eff_energy} 
except in the limit of perfectly coupled dust ($\tstop\to0$).  In
practice the dust-component would settle on a timescale
$t_\mathrm{settle}\sim 1/\OmK^2\tstop$, but for tightly-coupled dust
particles this is much longer than that to establish vertical
hydrostatic equilibrium in the gas component ($\sim 1/\OmK$).

True steady states may be calculated by invoking an underlying gas
turbulence, which leads to dust diffusion \citep{takeuchi02, youdin07, 
  lyra13}. This complication is beyond the scope of this paper and we 
defer it to a future work. Here, we will either consider $\tstop = 0$,
or assume that $\tstop$ is sufficiently small such that the background
disk (as obtained from Eq. \ref{steady_momr}---\ref{steady_momz}) does 
not evolve significantly.  %relative to perturbations

@@

\subsection{Vertical shear}\label{vertshear}
The mixture possess vertical shear. To see this, we eliminate $\Phi$
between Eq. \ref{steady_momr}---\ref{steady_momz} to 
obtain 
\begin{align}\label{vshear}
  r\frac{\p \Omega^2}{\p z} 
%&= \frac{\p\ln{\rho}}{\p r}\frac{\p}{\p
%    z}\left[c_s^2(1-\tepsilon)\right] - \frac{\p\ln{\rho}}{\p z}
%  \frac{\p}{\p r} \left[c_s^2(1-\tepsilon)\right]\\  
   = \frac{1}{\rho}\left(\frac{\p P}{\p r}\frac{\p \seff}{\p z} -\frac{\p
    P}{\p z}\frac{\p \seff}{\p r} \right). 
\end{align}
Writing Eq. \ref{vshear} in terms of the gas density and dust-to-gas
ratio with a power-law temperature profile (Eq. \ref{power_temp}) gives 
\begin{align}\label{vshear2}
  &r\frac{\p \Omega^2}{\p z}  =
  \frac{c_s^2(r)}{\left(1+\epsilon\right)^2}\left\{
  \frac{\p\epsilon}{\p r}\frac{\p\ln{\rhog}}{\p z}
%  -\frac{\p\epsilon}{\p z}\left[\frac{\p\ln{\rhog}}{\p r} + \frac{q}{r}\right]\right.\notag\\
 -\frac{\p\epsilon}{\p z}\frac{\p\ln{P}}{\p r}\right.\notag\\
  &\phantom{ r\frac{\p \Omega^2}{\p z}  =
    \frac{c_s^2(r)}{\left(1+\epsilon\right)^2}\left\{\right\} }
  \left. -\frac{q}{r} \left(1+\epsilon\right)\frac{\p\ln{\rhog}}{\p z} \right\} 
\end{align}
The first two terms correspond to vertical shear caused by spatial
variations in the dust-to-gas ratio; while the third term
proportional to $q$ corresponds to vertical shear due to the 
radial temperature gradient. The last term is present in the dust-free
limit.

@@

The average rate of work done is 
\begin{align}
  \mathcal{W} = \frac{1}{T_p}\int^{t+T_p}_{t}dt^\prime\int_M P
  \frac{Dv}{Dt^\prime} dm \label{work_def} 
\end{align}
\citep{cox67}, 
where $T_p$ is the oscillation period, $v=1/\rho$ is the specific
volume of the mixture, and the spatial integral is taken over the
total mass $M$ of the mixture and $dm$ is the mass element. 
Noting that only products of perturbations contribute to
$\mathcal{W}$, we find that for periodic oscillations (time dependence 
$e^{\ii\omega t}$ and real $\omega$): 
\begin{align}
  \mathcal{W} = - \frac{\omega}{2} \int \imag\left(\Delta P
  \frac{\Delta \rho^*}{\rho}\right)dV,\label{work_real}
\end{align}
where %the integral is taken over the volume of the
%mixture; and 
 the Lagragian perturbation $\Delta$ of a variable $f$ is 
$\Delta f = \delta f + \bm{\xi}\cdot\nabla f$; where $\bm{\xi}$ is the
Lagragian displacement ($    \xi_{x,z} =  \ii \dd v_{x,z}/\sigma$).  
Eq. \ref{work_real} show that a phase difference between gas pressure and
the total density leads to work done
($\mathcal{W}\neq0$).

Now, quite generally, Eq. \ref{lin_mass_full}---\ref{lin_energy_full} 
imply the integrand of the numerator in Eq. \ref{thermal_instability} 
is 
\begin{align} 
  \imag \left(\delta \mathcal{C}
  \nabla\cdot\dd \bm{v}^*\right) = 
  -|\sigma|^2\imag \left(\Delta P 
  \frac{\Delta\rho^*}{\rho}\right). \label{pdv}
\end{align}
%This states that `PdV' work (the right-hand-side) is 
%provided by dust-drag friction, and can be seen explicitly as 
%follows. 
Then combining Eq. \ref{work_real}, \ref{pdv} and
\ref{thermal_instability} 
gives  
\begin{align}
s = \frac{\mathcal{W}}{\mathcal{I}^2}, \label{growth_work}
\end{align}
where we have set $\sigma= - \omega$ in 
Eq. \ref{pdv}. Eq. \ref{growth_work} shows that if the average work
done is positive during a periodic oscillation, $\mathcal{W}>0$, then
then its amplitude would actually grow ($s>0$).

Under what condition is positive work is done? 
Consider a point in the volume where $\Delta\rho = 1$  
and take $\omega>0$ without loss of generality. (Thus the phase of
perturbations increase in time.) This point contributes to positive 
work done if $-\omega \imag\left(\Delta 
P\Delta\rho^*\right)>0$, which requires $\imag\left(\Delta
P\right)<0$. This corresponds to pressure lagging behind density since
$\imag\left(\Delta\rho\right)=0$.

@@

Fig. \ref{pdv_cartoon} gives a graphical demonstration that 
if gas pressure lags behind density, then positive work done 
($\mathcal{W}>0$) because it leads to a clockwise path in the
`P-V' plane. From $A$ to $B$, a fluid parcel is expanding to return to 
equilbrium, but pressure (i.e. restoring) forces is still
increasing. This over-compensation corresponds to positive work done. 
Similarly, from $C$ to $D$ the parcel is contracting but pressure 
(restoring) forces are still dropping, which allows the 
parcel's contraction to over-shoot its equilibrium density. The
overall positive work done leads to growth in the oscillation amplitude.

@@

When applied to dusty gas, the work done is attributed to finite 
dust-gas drag, which allows relative drift between gas and dust. 
A pressure-density lag of $A$ to $B$ is achieved if, 
after a region reaches maximum total density, its 
gas component is still increasing. This in turn implies a sufficiently
large particle flux \emph{out} of that region.

@@

This thermodynamic interpretation does not explain
\emph{why} dust drag causes pressure to lag behind density, but merely shows
that this \emph{must} be the case for any growing modes associated the
dust drag. To rigorously understand how dust-drag causes this
lag requires an explicit solution to the linearized equations with
detailed treatment of the function $\mathcal{C}$. However, given the complexity of
$\mathcal{C}$ (see Appendix \ref{lin_dust}), we might generally expect
the dusty disks to  support a range of stable and unstable modes, with the
latter being associated with pressure-density lag.

Indeed, \cite{jacquet11} explains the essence of the streaming 
instability in dusty protoplanetary disks as dust accumulation
(and hence compression) at a pressure bump, which then drags the 
gas towards it (and hence heating) to strengthen the pressure
bump. That is, heating occurs upon compression, as in stellar  
pulsational instabilities. In \S\ref{si} we check that the
streaming instability fits into this thermodynamic interpretation in
the strong drag limit.

>>>>

For a given distribution of the dust-fraction $\tepsilon$ (or
dust-to-gas ratio $\epsilon$), the 
mass and momentum Eqs. \ref{masseq}---\ref{momeq} admit     
axisymmetric steady states with $\rho(r,z)$ and 
$\bm{v}=r\Omega(r,z)\hat{\bm{\phi}}$ where $\Omega = v_\phi/r$, which satisfy 
\begin{align}
  r\Omega^2 &= \frac{\p \Phi}{\p r} + \frac{1}{\rho}\frac{\p P}{\p
    r},\label{steady_momr}\\
  0 & = \frac{\p\Phi}{\p z} + \frac{1}{\rho}\frac{\p P}{\p z},\label{steady_momz}
%  0 & = \nabla\cdot\left(\tepsilon\tstop\nabla P\right) \label{steady_dust}
\end{align}
with $P=P(\tepsilon,\rho)$ given by the equation of state
(Eq. \ref{eos}). An explicit solution is presented in
\S\ref{steady_state}.  
However, disk structures obtained this way generally do not satisfy 
the steady state effective energy Eq. \ref{eff_energy} because
$\mathcal{C}\neq0$ for general solutions to
Eq. \ref{steady_momr}---\ref{steady_momz}.

In practice the dust particles settle vertically on a timescale 
$t_\mathrm{settle}\sim 1/\OmK^2\tstop$, although for tightly-coupled dust
particles this is much longer than that to establish vertical
hydrostatic equilibrium in the gas component ($\sim 1/\OmK$).  
True steady states may be calculated by invoking an underlying gas
turbulence, which leads to dust diffusion \citep{takeuchi02, youdin07, 
 lyra13}. This complication is beyond the scope of this paper.

In this work, we will assume that $\tstop$  is sufficiently small so
that settling occurs on much longer timescales  than that of interest; or
consider equilibria with $\mathcal{C}\equiv 0$, namely   
\begin{inparaenum}[1)] 
\item 
  unstratified disks with finite dust-gas drag (\S\ref{si});  
\item 
  perfectly-coupled dust with $\tstop=0$ (\S\ref{results}). 
\end{inparaenum} 
Vertical settling is absent in both cases.

@@

\subsection{Vertical shear}\label{vertshear}
The mixture possess vertical shear. To see this, we eliminate $\Phi$
between Eq. \ref{steady_momr}---\ref{steady_momz} to 
obtain 
\begin{align}\label{vshear}
  r\frac{\p \Omega^2}{\p z} 
%&= \frac{\p\ln{\rho}}{\p r}\frac{\p}{\p
%    z}\left[c_s^2(1-\tepsilon)\right] - \frac{\p\ln{\rho}}{\p z}
%  \frac{\p}{\p r} \left[c_s^2(1-\tepsilon)\right]\\  
   = \frac{1}{\rho}\left(\frac{\p P}{\p r}\frac{\p \seff}{\p z} -\frac{\p
    P}{\p z}\frac{\p \seff}{\p r} \right). 
\end{align}
Writing Eq. \ref{vshear} in terms of the gas density and dust-to-gas
ratio with a power-law temperature profile (Eq. \ref{power_temp}) gives 
\begin{align}\label{vshear2}
  &r\frac{\p \Omega^2}{\p z}  =
  \frac{c_s^2(r)}{\left(1+\epsilon\right)^2}\left\{
  \frac{\p\epsilon}{\p r}\frac{\p\ln{\rhog}}{\p z}
%  -\frac{\p\epsilon}{\p z}\left[\frac{\p\ln{\rhog}}{\p r} + \frac{q}{r}\right]\right.\notag\\
 -\frac{\p\epsilon}{\p z}\frac{\p\ln{P}}{\p r}\right.\notag\\
  &\phantom{ r\frac{\p \Omega^2}{\p z}  =
    \frac{c_s^2(r)}{\left(1+\epsilon\right)^2}\left\{\right\} }
  \left. -\frac{q}{r} \left(1+\epsilon\right)\frac{\p\ln{\rhog}}{\p z} \right\} 
\end{align}
The first two terms correspond to vertical shear caused by spatial
variations in the dust-to-gas ratio; while the third term
proportional to $q$ corresponds to vertical shear due to the 
radial temperature gradient. The last term survives in the dust-free
limit.

@@

The average rate of work done is 
\begin{align}
  \mathcal{W} = \frac{1}{T_p}\int^{t+T_p}_{t}dt^\prime\int_M P
  \frac{D\upsilon}{Dt^\prime} dm \label{work_def} 
\end{align}
\citep{cox67}, 
where $T_p$ is the oscillation period, $\upsilon=1/\rho$ is the specific
volume of the mixture, and the spatial integral is taken over the
total mass $M$ of the mixture and $dm$ is the mass element. 
Noting that only products of perturbations contribute to
$\mathcal{W}$, we find that for periodic oscillations (time dependence 
$e^{\ii\omega t}$ and real $\omega$): 
\begin{align}
  \mathcal{W} = - \frac{\omega}{2} \int \imag\left(\Delta P
  \frac{\Delta \rho^*}{\rho}\right)dV,\label{work_real}
\end{align}
where %the integral is taken over the volume of the
%mixture; and 
 the Lagragian perturbation $\Delta$ of a variable $f$ is 
$\Delta f = \delta f + \bm{\xi}\cdot\nabla f$; where $\bm{\xi}$ is the
Lagragian displacement ($    \xi_{x,z} =  \ii \dd v_{x,z}/\sigma$).  
Eq. \ref{work_real} show that a phase difference between gas pressure and
the total density leads to work done
($\mathcal{W}\neq0$).

Now, quite generally, Eq. \ref{lin_mass_full}---\ref{lin_energy_full} 
imply the integrand of the numerator in Eq. \ref{thermal_instability} 
is 
\begin{align} 
  \imag \left(\delta \mathcal{C}
  \nabla\cdot\dd \bm{v}^*\right) = 
  -|\sigma|^2\imag \left(\Delta P 
  \frac{\Delta\rho^*}{\rho}\right). \label{pdv}
\end{align}
%This states that `PdV' work (the right-hand-side) is 
%provided by dust-drag friction, and can be seen explicitly as 
%follows. 
Then combining Eq. \ref{work_real}, \ref{pdv} and
\ref{thermal_instability} 
gives  
\begin{align}
s = \frac{\mathcal{W}}{\mathcal{I}^2}, \label{growth_work}
\end{align}
where we have set $\sigma= - \omega$ in 
Eq. \ref{pdv}. Eq. \ref{growth_work} states that if the average work
done is positive during a periodic oscillation, $\mathcal{W}>0$, then
then its amplitude would actually grow ($s>0$).  %and not be periodic

Positive work done is done by a parcel if $-\omega \imag\left(\Delta  
P\Delta\rho^*\right)>0$. Without loss of generality, take $\omega>0$
and consider a mass element with $\Delta\rho = 1$. Then positive work 
requires $\imag\left(\Delta
P\right)<0$.  This corresponds to pressure perturbations \emph{lagging}
behind that in density (or displacements), since the phase of
the assumed perturbations increase in time.

Fig. \ref{pdv_cartoon} gives a graphical demonstration that 
if gas pressure lags behind density, then %positive work done 
$\mathcal{W}>0$ because it leads to a clockwise path in the
`P-V' plane. From $A$ to $B$, a fluid parcel is expanding to return to 
equilbrium, but pressure (i.e. restoring) forces is still
increasing. This over-compensation causes the parcel to expand
beyond its equilibrium volume. 
%corresponds to positive work done. 
Similarly, from $C$ to $D$ the parcel is already contracting towards
equilibrium, but pressure
forces are still dropping. This allows the  
parcel's contraction to over-shoot its equilibrium density. The
overall positive work done leads to growth in the oscillation amplitude.

%$\imag\left(\Delta\rho\right)=0$.

@@

@@

%Applying the above argument to isothermal dusty gas, 
For dusty gas, the work done during oscillations 
 is attributed to finite dust-gas drag. The relative
drift between gas and dust causes a phase difference between the two
components.

A parcel of the isothermal dusty-gas mixture does 
postive work if  
\begin{align*}
-\sgn\left(\omega\right)\imag\left(\Delta\rhog\Delta\rhod^*\right)>0,
\end{align*}
meaning that gas follows dust. Instabilities are thus not possible if the
gas does not respond to dust-drag (i.e. no back-reaction). The
pressure-density lag shown in Fig. \ref{pdv_cartoon} is achieved if, 
after a parcel reaches maximum total density, its 
gas content is still increasing ($A$ to $B$). This in turn implies a sufficiently
large particle flux \emph{out} of the parcel.

{\bf lagrangian derivative follows the mixture. this highlights one
  perk of the one-fluid framework. in two-fluid models there are two
  types of lagragian deriv, not clear which fluid to follow.}

@@

This thermodynamic interpretation does not explain 
\emph{why} drag forces causes gas pressure to lag behind the dust
density, but merely shows 
that this \emph{must} be the case for any growing modes associated the
dust-gas drag. To rigorously understand how dust-drag causes this 
lag requires an explicit solution to the linearized equations with 
detailed treatment of the function $\mathcal{C}$. However, given the complexity of
$\mathcal{C}$ (see Appendix \ref{lin_dust}), we might generally expect
the dusty disks to  support a range of stable and unstable modes, with the
latter being associated with pressure-density lag.

Indeed, \cite{jacquet11} explains the essence of the streaming 
instability in dusty protoplanetary disks as dust accumulation
(and hence compression) at a pressure bump, which then drags the 
gas towards it (and hence heating) to strengthen the pressure
bump. That is, heating occurs upon compression, as in stellar  
pulsational instabilities \citep{cox67}. In \S\ref{si} we check that the
streaming instability fits into this thermodynamic interpretation in
the strong drag limit.
